# Setup
```{r}
rm(list = ls())
gc()

library(devtools)
if (!require("ComBatFamQC", quietly = TRUE)) {
  install_github("Zheng206/ComBatFamQC", build_vignettes = TRUE)
}

library(ComBatFamQC)
library(ggplot2)
library(dplyr)

setwd("/Users/advaymonga/Desktop/STARAPTOR")
```

# Example Vignettes
```{r}
vignette("ComBatQC")
vignette("Post-Harmonization")
```

# Load Data
```{r}
df_u <- read.csv("data/u_image_features.csv")
df_c <- read.csv("data/c_image_features.csv")
```

# Prepare Data for Harmonization
```{r}
feature_cols <- setdiff(
  colnames(df_u), 
  c("Subject_ID", "eGFR_12M", "DGF")
)
covariates <- c("eGFR_12M", "DGF")

df_u_prep <- df_u[, c("Subject_ID", feature_cols, 
                       "eGFR_12M", "DGF")]
df_c_prep <- df_c[, c("Subject_ID", feature_cols, 
                       "eGFR_12M", "DGF")]

df_u_prep$batch <- "U"
df_c_prep$batch <- "C"
combat_input <- rbind(df_u_prep, df_c_prep)

cat(" Dimensions: ", dim(combat_input), "\n")
```

# Batch Effect Diagnostics
```{r}
batch_diagnostics <- visual_prep(
  df = combat_input,
  type = "lm",
  features = feature_cols,
  batch = "batch",
  covariates = covariates
)

comfam_shiny(batch_diagnostics)
```

# ComBat Harmonization
```{r}
combat_result <- combat_harm(
  df = combat_input,
  features = feature_cols,
  batch = "batch",
  covariates = covariates,
  type = "lm",
  eb = TRUE,
  quiet = TRUE
)

harmonized_df <- combat_result$harmonized_df

gc(verbose = FALSE)
```

```{r}
# Save files
dir.create("data/harmonized_combined", 
           showWarnings = FALSE, recursive = TRUE)
dir.create("data/harmonized_u", 
           showWarnings = FALSE, recursive = TRUE)
dir.create("data/harmonized_c", 
           showWarnings = FALSE, recursive = TRUE)

harmonized_u <- harmonized_df[
  harmonized_df$batch == "U",
  !(names(harmonized_df) %in% "batch")
]
harmonized_c <- harmonized_df[
  harmonized_df$batch == "C",
  !(names(harmonized_df) %in% "batch")
]

write.csv(harmonized_df, 
          "data/harmonized_combined/harmonized_all.csv", 
          row.names = FALSE)

write.csv(harmonized_u, 
          "data/harmonized_u/u_harmonized.csv", 
          row.names = FALSE)
write.csv(harmonized_c, 
          "data/harmonized_c/c_harmonized.csv", 
          row.names = FALSE)

rm(harmonized_u, harmonized_c)
gc(verbose = FALSE)
```

# Post-Harmonization Diagnostics
```{r}
post_diagnostics <- visual_prep(
  df = harmonized_df,
  type = "lm",
  features = feature_cols,
  batch = "batch",
  covariates = covariates
)

comfam_shiny(post_diagnostics)
```

# CovBat Harmonization
```{r}
cat("Running CovBat harmonization...\n")

covbat_result <- combat_harm(
  type = "lm",
  features = feature_cols,
  batch = "batch",
  covariates = covariates,
  df = combat_input,
  family = "covfam",
  eb = TRUE,
  quiet = TRUE
)

covbat_harmonized_df <- covbat_result$harmonized_df

gc(verbose = FALSE)
cat("CovBat harmonization complete.\n")
```

# Save CovBat Results
```{r}
dir.create("data/covbat_harmonized_combined", 
           showWarnings = FALSE, recursive = TRUE)
dir.create("data/covbat_harmonized_u", 
           showWarnings = FALSE, recursive = TRUE)
dir.create("data/covbat_harmonized_c", 
           showWarnings = FALSE, recursive = TRUE)

covbat_harmonized_u <- covbat_harmonized_df[
  covbat_harmonized_df$batch == "U",
  !(names(covbat_harmonized_df) %in% "batch")
]
covbat_harmonized_c <- covbat_harmonized_df[
  covbat_harmonized_df$batch == "C",
  !(names(covbat_harmonized_df) %in% "batch")
]

write.csv(covbat_harmonized_df, 
          "data/covbat_harmonized_combined/covbat_harmonized_all.csv", 
          row.names = FALSE)
write.csv(covbat_harmonized_u, 
          "data/covbat_harmonized_u/u_covbat_harmonized.csv", 
          row.names = FALSE)
write.csv(covbat_harmonized_c, 
          "data/covbat_harmonized_c/c_covbat_harmonized.csv", 
          row.names = FALSE)

rm(covbat_harmonized_u, covbat_harmonized_c)
gc(verbose = FALSE)

cat("CovBat results saved.\n")
```

# CovBat Post-Harmonization Diagnostics
```{r}
covbat_post_diagnostics <- visual_prep(
  df = covbat_harmonized_df,
  type = "lm",
  features = feature_cols,
  batch = "batch",
  covariates = covariates
)

comfam_shiny(covbat_post_diagnostics)
```

# ─────────────────────────────────────────────────────────────
# LEAVE-ONE-COHORT-OUT (LOO) COMBAT
# ─────────────────────────────────────────────────────────────
#
# For each of the 3 combinations:
#   1. Fit ComBat on the 2 training cohorts (full EB, covariate-adjusted)
#   2. Apply the learned model to the held-out test cohort via predict()
#      — no ComBat re-fitting on the test data
#   3. Save harmonized train and test CSVs for 03_loo_combat.py
#
# Combinations:
#   UC → M : train on UC Davis + Coimbra,  test on Mayo
#   UM → C : train on UC Davis + Mayo,     test on Coimbra
#   CM → U : train on Coimbra  + Mayo,     test on UC Davis

# Load Mayo data (not loaded earlier in this file)
```{r}
df_m <- read.csv("data/m_image_features.csv")

# feature_cols and covariates already defined above
# Confirm M has the same features
cat("M cohort dimensions:", dim(df_m), "\n")
cat("Feature columns match U:", all(feature_cols %in% colnames(df_m)), "\n")
```

# LOO Helper: fit ComBat on training cohorts, apply to test cohort
```{r}
run_loo_combat <- function(train_list, train_labels, df_test, test_label,
                           feature_cols, covariates, out_dir) {

  # --- Build training data frame with batch labels ---
  train_dfs <- mapply(function(df, lbl) {
    df_prep        <- df[, c("Subject_ID", feature_cols, covariates)]
    df_prep$batch  <- lbl
    df_prep
  }, train_list, train_labels, SIMPLIFY = FALSE)

  df_train_combined <- do.call(rbind, train_dfs)
  cat("  Training subjects:", nrow(df_train_combined),
      "(", paste(train_labels, collapse = "+"), ")\n")

  # --- Step 2: Fit ComBat on training cohorts ---
  combat_fit <- combat_harm(
    df         = df_train_combined,
    features   = feature_cols,
    batch      = "batch",
    covariates = covariates,
    type       = "lm",
    eb         = TRUE,
    quiet      = TRUE
  )

  harmonized_train <- combat_fit$harmonized_df
  cat("  ComBat fit complete.\n")

  # --- Step 4: Apply learned parameters to test cohort (no refitting) ---
  df_test_prep        <- df_test[, c("Subject_ID", feature_cols, covariates)]
  df_test_prep$batch  <- test_label

  # predict() uses the fitted ComBat model to estimate and remove the test
  # cohort's batch effect — the model is not re-fitted on the test data
  harmonized_test_features <- predict(
    combat_fit$ComBat.model,
    newdata = df_test_prep,
    batch   = "batch"
  )

  # Reassemble with metadata
  harmonized_test <- df_test_prep
  harmonized_test[, feature_cols] <- harmonized_test_features[, feature_cols]

  cat("  Test cohort harmonized:", nrow(harmonized_test), "subjects (", test_label, ")\n")

  # --- Save ---
  dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

  # Drop batch column before saving (Python pipeline doesn't need it)
  train_out <- harmonized_train[, !(names(harmonized_train) %in% "batch")]
  test_out  <- harmonized_test[,  !(names(harmonized_test)  %in% "batch")]

  write.csv(train_out, file.path(out_dir, "train_harmonized.csv"), row.names = FALSE)
  write.csv(test_out,  file.path(out_dir, "test_harmonized.csv"),  row.names = FALSE)

  cat("  Saved to", out_dir, "\n")
  invisible(list(train = harmonized_train, test = harmonized_test))
}
```

# LOO Fold 1: Train UC Davis + Coimbra → Test Mayo
```{r}
cat("=" , strrep("=", 59), "\n")
cat("LOO Fold 1: Train UC Davis + Coimbra  →  Test Mayo\n")
cat("=" , strrep("=", 59), "\n")

loo_UC_to_M <- run_loo_combat(
  train_list   = list(df_u, df_c),
  train_labels = c("U", "C"),
  df_test      = df_m,
  test_label   = "M",
  feature_cols = feature_cols,
  covariates   = covariates,
  out_dir      = "data/loo_combat/UC_to_M"
)

gc(verbose = FALSE)
```

# LOO Fold 2: Train UC Davis + Mayo → Test Coimbra
```{r}
cat("=" , strrep("=", 59), "\n")
cat("LOO Fold 2: Train UC Davis + Mayo  →  Test Coimbra\n")
cat("=" , strrep("=", 59), "\n")

loo_UM_to_C <- run_loo_combat(
  train_list   = list(df_u, df_m),
  train_labels = c("U", "M"),
  df_test      = df_c,
  test_label   = "C",
  feature_cols = feature_cols,
  covariates   = covariates,
  out_dir      = "data/loo_combat/UM_to_C"
)

gc(verbose = FALSE)
```

# LOO Fold 3: Train Coimbra + Mayo → Test UC Davis
```{r}
cat("=" , strrep("=", 59), "\n")
cat("LOO Fold 3: Train Coimbra + Mayo  →  Test UC Davis\n")
cat("=" , strrep("=", 59), "\n")

loo_CM_to_U <- run_loo_combat(
  train_list   = list(df_c, df_m),
  train_labels = c("C", "M"),
  df_test      = df_u,
  test_label   = "U",
  feature_cols = feature_cols,
  covariates   = covariates,
  out_dir      = "data/loo_combat/CM_to_U"
)

gc(verbose = FALSE)
cat("\nLOO ComBat harmonization complete.\n")
cat("Run 03_loo_combat.py to train and evaluate models.\n")
```
